A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-10 paper/paper.Rmd
DC completed the first two of her changes

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].

@@

This disparity between the average and the individual is a part of statistics as a discipline that requires more attention. This particular data set is a prime example of discussing this disparity. Textbook data sets have longevity if there is an unresolved mystery. The iris data [@iris-data] is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]). We argue the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

@@

A robust linear regression model using the `rlm` function from `MASS` package [@mass] is used to treat the extreme values in the data. The robustness weight is used to determine if a value should be replaced with the fitted value from the model. This is constructed for each ID utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively. An alternative approach would be a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP]. However, when we tested this, the fit failed to converge. <!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.

The `mean_hourly_wage` and `year` are set as the dependent and predictor, respectively. Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm]. The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. It should be noted that it is not possible to determine if all outliers are errors, and it might be an individual indeed had abnormally high wages at a particular time.

To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done. Samples of individuals were examined under the different thresholds with an eye to the smoothness of the profiles. Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals. A threshold of 0.12 was chosen. That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called `is_pred`, so that this change can be monitored in the downstream analyses.

@@

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.

@@

The data cleaning process is documented, and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a `shiny`  [@shiny] app, and the choices used in the refreshed data are documented.

@@

Why is this exercise important for teachers and students? This work illustrates the steps in cleaning and processing data in preparation for it to be a textbook data set. Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent. Along with the textbook data, which is now provided as an R package, the documentation of the process provides some examples for teaching data cleaning, initial data analysis, and exploratory data analysis. The refreshed textbook data provides a resource for teaching longitudinal data analysis.

@@

- **Codes**: R script to reproduce data tidying and cleaning is available in this [page](`r yowie_pkgdown`/articles/process-data.html).

>>>>

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].

@@

This disparity between the average and the individual is a part of statistics as a discipline that requires more attention. This particular data set is a prime example of discussing this disparity. Textbook data sets have longevity if there is an unresolved mystery. The iris data [@iris-data] is a prime example. It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]). We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?

@@

A robust linear regression model using the `rlm` function from `MASS` package [@mass] is used to treat the extreme values in the data. The robustness weight is used to determine if a value should be replaced with the fitted value from the model. This is constructed for each ID utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively. An alternative approach would be a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP]. <!-- However, when we tested this, the fit failed to converge. I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->

Figure \ref{fig:compare-plot} shows the profiles for a sample of 30 individuals the mean hourly wage before and after extreme values are replaced. The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.  <!--The `mean_hourly_wage` and `year` are set as the dependent and predictor, respectively. Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm]. -->

The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers. It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time. To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed. Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined. Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals. This was achieved using a purpose written `shiny`  [@shiny] app (provided with the code of this paper). <!--A threshold of 0.12 was chosen. That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.--> Observations where a change was made are tracked with a new variable called `is_pred`, so that this effect in the downstream analyses can be monitored.

@@

@@

The data cleaning process is documented, and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database. Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a `shiny`  [@shiny] app, which is provided with the code, and the choices used in the refreshed data are documented.

@@

Why is the work presented here useful for teachers and students? There are several reasons:

- It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges. Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
- Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
- The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current in time.
- It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience. The vignettes of the `brolgar` [@brolgar] package provide useful guides for the case study content.

@@

- **Codes**: R script to reproduce data tidying and cleaning is available in this [page](`r yowie_pkgdown`/articles/process-data.html). The code for the extreme value handling with robust linear fits is in \url{`r yowie_pkgdown`/articles/input-to-valid-data.html}.
